---
title: "Moving spatial proteomics data between pRoloc and grassp"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Moving spatial proteomics data between pRoloc and grassp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why this package exists

[pRoloc](https://bioconductor.org/packages/pRoloc/) and
[grassp](https://github.com/czbiohub-sf/grassp) are the R and Python frameworks for the same
science: inferring protein subcellular localisation from fractionation proteomics. Their data
models are nearly the same shape — a features-by-fractions quantitation matrix with feature
and sample metadata — but they share no file format, so an analysis has historically had to
pick one language and stay there.

`grasspio` supplies both directions of the conversion, using **h5ad** as the exchange format.
That lets you preprocess in one language and classify in the other. It is also useful on its
own: neither pRoloc nor MSnbase ships an exporter, so `grassp_write_msnset()` is a way to hand
a classified `MSnSet` to anything that reads h5ad.

This vignette is about moving *your own* object across the boundary. If instead you want to pull
one of the ~100 curated datasets off the
[grassp data portal](https://grassp.apps.czbiohub.org/datasets) and analyse it in pRoloc, see
`vignette("portal", package = "grasspio")` — that needs no Python either.

## Installation

```{r install}
install.packages("remotes")

# rhdf5 first: anndataR calls it but declares it only in Suggests, so nothing else pulls it in.
BiocManager::install("rhdf5")

# anndataR itself is a hard dependency of grasspio and arrives with it. It is scverse's native
# R implementation of the AnnData on-disk spec, so no Python environment is involved.
remotes::install_github("czbiohub-sf/grassp", subdir = "r/grasspio")
```

grasspio needs R >= 4.5, inherited from anndataR.

## Starting from Python

Someone hands you an `experiment.h5ad` produced by grassp's `gr.io.write_msnset()`. The file
is self-describing, so reading it takes no arguments:

```{r read}
library(grasspio)
library(pRoloc)

x <- grassp_as_msnset("experiment.h5ad")
x
fvarLabels(x)                            # what marker columns are available
getMarkerClasses(x, fcol = "markers")    # pick one per call
```

Note that nothing designates a single marker column. `fcol` is a per-call argument, so an
object can carry `markers`, `markers.orig`, `pd.markers` and more at once, and you point each
function at whichever you want — the same freedom AnnData gives you.

From here it is an ordinary `MSnSet`, so the whole pRoloc workflow applies:

```{r classify}
params <- svmOptimisation(x, fcol = "markers", times = 100, xval = 5,
                          class.weights = classWeights(x, fcol = "markers"))
x <- svmClassification(x, params, fcol = "markers", scores = "all")

ts <- orgQuants(x, fcol = "svm", scol = "svm.scores", t = 0.75)
x <- getPredictions(x, fcol = "svm", scol = "svm.scores", t = ts)

x <- tagmMapPredict(x, params = tagmMapTrain(x, fcol = "markers"), probJoint = TRUE)

plot2D(x, fcol = "svm.pred")
addLegend(x, fcol = "svm.pred", where = "topleft")
```

And send the results back:

```{r write}
grassp_write_msnset(x, "results.h5ad")
```

The Python side reads that straight back into an AnnData:

```python
annotated = gr.io.read_msnset("results.h5ad")
```

and gets back everything it sent, plus what pRoloc added. If it would rather keep working on the
object already in its session, `gr.io.read_proloc_results("results.h5ad", adata)` grafts the new
`obs` columns and `obsm` matrices onto that instead.

## Starting from R

The same functions work if R is where your data lives. Export your own `MSnSet` and a
collaborator can pick it up in Python:

```{r r-first}
library(pRolocdata)
data(dunkley2006)

grassp_write_msnset(dunkley2006, "dunkley2006.h5ad")
```

## What crosses over

| `MSnSet` | AnnData |
| --- | --- |
| `exprs()` | `X` — **no transpose**; grassp keeps proteins in `obs`, matching the features-by-fractions orientation |
| `featureNames()` / `sampleNames()` | `obs_names` / `var_names` |
| `fData()` scalar columns | `obs` columns |
| `fData()` **matrix-valued** columns | `obsm`, with names in `uns$obsm_colnames` |
| `pData()` | `var` |
| `pData()` **matrix-valued** columns | `varm`, with names in `uns$varm_colnames` |
| extra `assayData` elements | `layers` |
| `processingData()@processing` | `uns$processing` |
| `experimentData()@other$grassp_uns` | everything else in `uns` |
| `"unknown"` | `NA` |

The matrix-column row is the one that makes this work. pRoloc stores per-protein by
per-compartment score matrices *inside a single `fData` column* — `svm.all.scores`,
`tagm.map.joint`, `tagm.mcmc.joint`, `bandle.joint`, `Markers`, `GOAnnotations` — and that is
exactly what AnnData's `obsm` is for. Since `obsm` arrays have no column names, the class
names are carried alongside in `uns$obsm_colnames`:

```{r matrix}
x <- grassp_as_msnset("results.h5ad")
is.matrix(fData(x)$svm.all.scores)
colnames(fData(x)$svm.all.scores)
```

The `experimentData@other` row is what makes a grassp → pRoloc → grassp trip lossless rather
than merely useful. An `MSnSet` has no slot for arbitrary metadata, so `grassp_as_msnset()` parks
whatever `uns` entries the contract does not map — neighbour parameters, PCA variance ratios,
colour maps — on `experimentData(x)@other$grassp_uns`, and `grassp_write_msnset()` hands them
straight back. pRoloc never looks there. It is safe here for the reason it would not be for a
graph: `uns` is neither row- nor column-aligned, so subsetting has nothing to get wrong.

The contract's own `uns` keys (`msnset_spec`, `msnset_dropped`, `obsm_colnames`,
`varm_colnames`, ...) are the exception — they are regenerated from the object on every write, so
an artifact describes what is actually in it even after you have subset in R.

Two conversions are handled for you, and both are easy to get wrong by hand — which is the
main argument against rolling your own CSV:

- **`"unknown"` ↔ `NA`.** pRoloc encodes unlabelled features as the literal string
  `"unknown"`; every classifier picks its training set with
  `fData(object)[, fcol] != "unknown"`, and `markerMSnSet()`/`unknownMSnSet()` fail outright on
  `NA`. grassp uses `NA` instead. Get this wrong and you quietly train a compartment class
  called "unknown". Both directions convert every text column, not a nominated one.
- **Class names matched by name, never by position.** R sorts factor levels, and
  `getMarkerClasses()` sorts, so a probability matrix does not come back in the column order
  it left in.

## Limitations

- AnnData `obsp`/`varp` do not cross into R, and are the only slots that do not: `eSet` has no
  pairwise slot, and pRoloc has no graph structure to map onto. They are listed in
  `uns$msnset_dropped`. Little is lost — a graph is derived from the quantitation matrix, so the
  Python side rebuilds one with a single `gr.pp.neighbors()` call, and `plot2D()` recomputes its
  own projection regardless.
- Layers arrive as extra `assayData` elements and `varm` as matrix `pData` columns. Both subset
  correctly with `[`, `markerMSnSet()` and `filterNA()`, and `exprs()` stays the matrix pRoloc's
  functions use:

  ```{r extra}
  assayDataElementNames(x)          # "exprs", "pvals", ...
  assayDataElement(x, "pvals")
  is.matrix(pData(x)$PCs)           # a varm entry
  ```
- Multi-localisation travels as the binary `Markers` matrix, like any other matrix `fData`
  column. Note that `mrkMatToVec()` collapses any protein with more than one label to
  `"unknown"`, and that pRoloc's classifiers cannot train on the matrix encoding — it is for
  inspecting and subsetting marker sets (`getMarkerClasses`, `markerMSnSet`, `unknownMSnSet`).
- Result objects that live outside `fData()` — `GenRegRes` from `svmOptimisation()`,
  `MAPParams`, `MCMCParams`, `bandleParams`, `QSep` — do not cross over. Only the summaries
  pRoloc writes into `fData()` do. Save those separately with `saveRDS()` if you need them;
  re-running MCMC to recover one is expensive.

## Contract versioning

Artifacts record the contract version they were written with, and both sides check it, so a
file from a newer grassp fails loudly rather than being misread:

```{r spec}
grassp_msnset_spec()
#> [1] "grassp-msnset/1"
```
