---
title: "Working with grassp portal datasets in pRoloc"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Working with grassp portal datasets in pRoloc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

# Working with grassp portal datasets in pRoloc

The [grassp data portal](https://grassp.apps.czbiohub.org/datasets) publishes curated
subcellular fractionation datasets — around a hundred of them, uniformly processed, each with
several published marker sets already mapped onto it. They are stored as `.h5ad`, which sounds
like a Python format, but `grasspio` reads it natively in R.

So this vignette needs **no Python at all**. You download a file, get an `MSnSet`, and
from there it is ordinary pRoloc.

The other vignette, `vignette("grasspio")`, is about the symmetric exchange format: taking your
*own* object between R and Python. This one is about consuming the portal.

## Installation

```{r install, eval = FALSE}
install.packages(c("remotes", "BiocManager"))

# rhdf5 first. anndataR needs it to touch an HDF5 file but declares it only in Suggests, so
# nothing else pulls it in.
BiocManager::install(c("rhdf5", "pRoloc"))

# grasspio depends on anndataR outright, so that arrives with it.
remotes::install_github("czbiohub-sf/grassp", subdir = "r/grasspio")
```

Needs R >= 4.5, which is anndataR's own requirement. `pRoloc` is only needed to *analyse* the
result — `grasspio` itself can read a portal file without it.

```{r libs, message = FALSE, warning = FALSE}
library(grasspio)
library(pRoloc)
library(MSnbase)
```

```{r can-run, include = FALSE}
# Everything below needs the network. Rather than failing `R CMD check` on an offline machine,
# probe once and let the remaining chunks no-op. If you are reading a version of this page with
# no output, that is what happened.
DATASET <- "Currie_2024_AC16_Control"
PORTAL <- "https://public.czbiohub.org/proteinxlocation/datasets/"
options(timeout = max(600, getOption("timeout")))
h5ad <- file.path(tempdir(), paste0(DATASET, ".h5ad"))
CAN_RUN <- tryCatch(
  {
    download.file(paste0(PORTAL, DATASET, ".h5ad"), h5ad, mode = "wb", quiet = TRUE)
    file.exists(h5ad) && file.size(h5ad) > 1e6
  },
  error = function(e) FALSE, warning = function(w) FALSE
)
knitr::opts_chunk$set(eval = CAN_RUN)
```

## Getting a dataset

Browse [the portal](https://grassp.apps.czbiohub.org/datasets), pick a dataset, and take its
`.h5ad`. The files sit at a predictable URL, so you can either download one in your browser or
fetch it from R:

```
https://public.czbiohub.org/proteinxlocation/datasets/<name>.h5ad       # processed
https://public.czbiohub.org/proteinxlocation/datasets_raw/<name>.h5ad   # pre-enrichment
```

The `datasets/` objects are what you usually want: enriched, replicates collapsed, and each
protein's profile already normalised to sum to 1. The `datasets_raw/` objects are raw
intensities with replicates preserved — useful, but they need normalising before pRoloc's
distance-based methods mean anything.

Here we take one example file, `Currie_2024_AC16_Control` — a LOPIT-DC experiment on AC16
cardiomyocytes — with `download.file()`:

```{r download, eval = FALSE}
options(timeout = 600) # the default 60 s is not enough for the larger objects
h5ad <- "Currie_2024_AC16_Control.h5ad"
download.file(
  paste0("https://public.czbiohub.org/proteinxlocation/datasets/", h5ad),
  h5ad,
  mode = "wb"
)
```

and read it straight into an `MSnSet`:

```{r read}
x <- grassp_as_msnset(h5ad)
x
```

That is the whole conversion. A portal file is a *plain* h5ad — it carries none of `grasspio`'s
own exchange-contract metadata — and `grassp_as_msnset()` reads it anyway: the contract check
tolerates a missing version block, and the `nan_to_unknown` default rewrites `NA` markers to the
literal `"unknown"` that pRoloc requires (`markerMSnSet()` fails outright on `NA`).

## Getting oriented

Everything in the AnnData object has landed somewhere. Per-protein annotations are `fData`,
per-fraction ones are `pData`, and the extra matrices are additional `assayData` elements:

```{r orient}
dim(x)
head(fvarLabels(x), 12)
assayDataElementNames(x)
pData(x)[, c("fraction", "fraction_order")]
```

Some `fData` columns are themselves **matrices** — the per-protein × per-compartment arrays that
came from AnnData's `obsm`. That is pRoloc's own idiom for score matrices, so they arrive in the
shape pRoloc expects:

```{r matrix-columns}
vapply(fData(x), function(col) if (is.matrix(col)) paste(dim(col), collapse = " x ") else "",
       character(1))[c("X_pca", "X_umap", "harmonized_annotation_propagated_probabilities")]
```

The publication metadata came too. An `MSnSet` has no slot for arbitrary key-value metadata, so
`grasspio` parks whatever it cannot map on `experimentData(x)@other$grassp_uns`:

```{r provenance}
uns <- experimentData(x)@other$grassp_uns
unlist(uns[c("title", "publication_journal", "publication_doi")])
```

### Pick a marker column — there is no single one

The portal maps several published marker sets onto every dataset. pRoloc's `fcol` is a per-call
argument, so all of them are usable and you choose per analysis:

```{r markers}
marker_cols <- grep("^marker_|^author_markers", fvarLabels(x), value = TRUE)
data.frame(
  column  = marker_cols,
  classes = vapply(marker_cols, function(f) length(getMarkerClasses(x, fcol = f)), integer(1)),
  labelled = vapply(marker_cols, function(f) sum(fData(x)[[f]] != "unknown"), integer(1)),
  row.names = NULL
)
```

We will use `marker_lilley`, the set `pRoloc::pRolocmarkers()` also ships:

```{r fcol}
FCOL <- "marker_lilley"
getMarkerClasses(x, fcol = FCOL)
dim(markerMSnSet(x, fcol = FCOL))
```

```{r no-prep}
# No preprocessing needed on an enriched object: filterNA() and normalise() would be no-ops.
c(missing = sum(is.na(exprs(x))), rows_summing_to_1 = sum(abs(rowSums(exprs(x)) - 1) < 1e-6))
```

**One trap before you plot.** pRoloc defaults to `fcol = "markers"`, and a portal object has no
such column — so a bare `plot2D(x)` errors with
`fcol %in% fvarLabels(object) is not TRUE`. Always name your column, or pass `fcol = NULL`.

## Plotting

```{r plot2d-default, fig.alt = "PCA of the fractionation profiles coloured by marker class"}
plot2D(x, fcol = FCOL, main = "pRoloc default PCA (scaled)")
addLegend(x, fcol = FCOL, where = "topright", cex = 0.6, ncol = 2)
```

```{r scree, fig.alt = "Scree plot showing nine informative components", fig.height = 4}
plot2D(x, method = "scree")
```

Note the tenth bar is empty. The profiles are normalised to sum to 1, so the centred matrix is
rank 9 — which is also why the portal stores nine PCA components for ten fractions, not ten.

## Comparing pRoloc's PCA with the portal's own

The portal object already carries PCA coordinates in `fData(x)$X_pca`, computed by grassp when
the dataset was built. Recomputing them in pRoloc is a good way to check you are looking at the
same object — but only if you match the convention, and **pRoloc's default does not**:

```{r pca-default}
stored <- fData(x)$X_pca
colnames(stored) <- paste0("PC", seq_len(ncol(stored)))

default_pca <- plot2D(x, fcol = FCOL, plot = FALSE)
colnames(default_pca) # pRoloc prints the variance explained into the names
c(PC1 = cor(default_pca[, 1], stored[, 1]), PC2 = cor(default_pca[, 2], stored[, 2]))
```

PC2 correlates at about 0.24. Nothing is broken: `plot2D()` passes `scale = TRUE` to `prcomp`,
scaling every fraction to unit variance, while grassp centres without scaling. Scaling does not
merely stretch the axes — it changes the eigenvectors *and their order*, so comparing
component-by-component across the two conventions compares different directions.

Match the convention and they agree:

```{r pca-matched}
matched <- plot2D(x, fcol = FCOL, plot = FALSE,
                  methargs = list(center = TRUE, scale = FALSE))
colnames(matched)
signed_cor <- vapply(1:2, function(k) cor(matched[, k], stored[, k]), numeric(1))
signed_cor
```

The variance-explained figures in the column names are the quickest tell that you have the right
convention: 39.8% / 32.9% here versus 44.5% / 27.4% for the scaled default.

`cor(PC2) = -1` is not a disagreement either. An eigenvector is only defined up to sign, so some
components come back negated; that is arithmetic, not a discrepancy. Fix it explicitly:

```{r pca-signs}
recomputed <- prcomp(exprs(x), center = TRUE, scale. = FALSE)$x[, 1:ncol(stored)]
flip <- sign(diag(cor(recomputed, stored)))
round(flip, 0)
aligned <- sweep(recomputed, 2, flip, `*`)
round(diag(cor(aligned, stored)), 6)
```

Those correlations round to 1, but the coordinates themselves still differ by around `1e-4`:
the portal ships the embedding its pipeline produced, not a bit-exact function of the matrix
alongside it. Treat it as a fixed reference embedding — compare with a correlation or a
tolerance of about `1e-3`, not with `all.equal()`.

### Drawing grassp's embeddings with pRoloc

`plot2D(method = "none")` takes coordinates you already have and styles them like any other
pRoloc plot. That works for the stored PCA and, usefully, for the UMAP — which spares you
compiling a UMAP package just to look at it:

```{r plot-none, fig.show = "hold", fig.width = 4.2, fig.height = 4.2, fig.alt = "The portal's own stored PCA and UMAP coordinates, drawn in pRoloc's idiom"}
plot2D(stored[, 1:2], method = "none", methargs = list(x), fcol = FCOL,
       main = "portal X_pca")
umap <- fData(x)$X_umap
colnames(umap) <- c("UMAP1", "UMAP2")
plot2D(umap, method = "none", methargs = list(x), fcol = FCOL, main = "portal X_umap")
```

## Profiles and marker quality

`plotDist()` is most informative when you give it two things: the proteins *assigned* to a
compartment, and — via `markers =` — the subset of those that are actually markers for it. The
markers are drawn in colour over the rest in grey, so you can see at a glance whether the
assigned population really follows the marker profile.

The portal makes that easy because it carries both an author annotation and several published
marker sets. One catch: they do **not** share a vocabulary. `marker_lilley` says
`"Mitochondrion"` where the authors' own column says `"MITOCHONDRION"`, and `"Golgi"` is `"GA"`.
Nothing in the bridge translates compartment names — that is deliberate, since silently renaming
someone's labels is worse than a mismatch — so map them yourself:

```{r plotdist, fig.show = "hold", fig.width = 4.2, fig.height = 3.9, fig.alt = "Profiles of all proteins assigned to a compartment, with that compartment's markers highlighted"}
to_author <- c(Mitochondrion = "MITOCHONDRION", ER = "ER")

for (cl in names(to_author)) {
  assigned <- featureNames(x)[fData(x)$author_annotation == to_author[[cl]]]
  markers <- featureNames(x)[fData(x)[[FCOL]] == cl]
  # plotDist() subsets the object it is *given* by `markers`, so a marker the authors placed
  # somewhere else is out of bounds. Pass the intersection; the shortfall is itself worth
  # seeing, since it counts markers the two annotations disagree about.
  shown <- intersect(markers, assigned)
  # pch = 1 because plotDist() hands `...` to matplot(), which otherwise warns that it is
  # recycling the default pch across the ten fraction columns.
  plotDist(x[assigned, ], markers = shown, pch = 1,
           main = sprintf("%s\n%d assigned, %d of %d markers", cl,
                          length(assigned), length(shown), length(markers)))
}
```

```{r consprofiles, fig.alt = "Consensus profile heatmap across marker classes", fig.height = 4}
plotConsProfiles(mrkConsProfiles(x, fcol = FCOL))
```

`QSep()` quantifies how well separated the marker classes are — the single most useful sanity
check on a fractionation experiment, and something the portal makes unusually easy because you
can ask it of every marker set at once:

```{r qsep}
q <- QSep(x, fcol = FCOL)
summary(q)
```

```{r qsep-all}
sapply(marker_cols, function(f) median(summary(QSep(x, fcol = f), verbose = FALSE)))
```

Higher is better: it is the between-class distance in units of within-class spread, so a value
near 1 means a class is no further from its neighbours than from itself.

## Classification

`svmClassification()` needs hyperparameters. A real analysis gets them from
`svmOptimisation()`, which is a grid search over repeated stratified cross-validation and takes
minutes — so it is shown here but not run:

```{r svm-opt, eval = FALSE}
params <- svmOptimisation(x, fcol = FCOL, times = 100,
                          class.weights = classWeights(x, fcol = FCOL))
plot(params)
getParams(params)
```

```{r svm}
set.seed(1) # e1071's probability scaling consumes the RNG; without this, runs differ
x <- svmClassification(x, fcol = FCOL, sigma = 0.1, cost = 16, scores = "all")

# `scores = "all"` writes the per-class matrix but NOT the scalar score that orgQuants() and
# getPredictions() look for, so derive it.
fData(x)$svm.scores <- apply(fData(x)$svm.all.scores, 1, max)
dim(fData(x)$svm.all.scores)
```

Note the output columns are named `svm`, `svm.scores`, `svm.all.scores` regardless of which
`fcol` you trained on. (The bare `[1] "marker_lilley"` above the result is a stray `print()`
inside `MLInterfaces`, not something you did — every classification call emits one.)

```{r predictions}
# `mcol` is the second fcol-like trap: orgQuants() and getPredictions() default it to
# "markers" too, and use it to decide which proteins were training data.
ts <- orgQuants(x, fcol = "svm", scol = "svm.scores", mcol = FCOL, t = 0.75, verbose = FALSE)
ts[is.na(ts)] <- Inf # a class with too few markers has no quantile
x <- getPredictions(x, fcol = "svm", scol = "svm.scores", mcol = FCOL, t = ts,
                    verbose = FALSE)
table(fData(x)$svm.pred == "unknown")
```

```{r plot-svm, fig.alt = "PCA coloured by thresholded SVM prediction"}
plot2D(x, fcol = "svm.pred", methargs = list(center = TRUE, scale = FALSE),
       main = "SVM predictions, thresholded")
```

A second classifier is a cheap sanity check:

```{r knn}
set.seed(1)
x <- knnClassification(x, fcol = FCOL, k = 5) # note: knnClassification rejects `verbose`
mean(fData(x)$svm == fData(x)$knn)
```

And `QSep` closes the loop quantitatively — do the predicted labels separate better than the
markers you started from?

```{r qsep-pred}
c(markers = median(summary(QSep(x, fcol = FCOL), verbose = FALSE)),
  svm.pred = median(summary(QSep(x, fcol = "svm.pred"), verbose = FALSE)))
```

## Sending the result somewhere else

`grasspio` writes back out too, which is useful even if you never touch Python. Neither pRoloc
nor MSnbase ships an h5ad exporter, so this is also how you would package a pRoloc-processed
dataset for submission to the [grassp portal](https://grassp.apps.czbiohub.org/datasets):

```{r write}
out <- file.path(tempdir(), "currie_classified.h5ad")
grassp_write_msnset(x, out, overwrite = TRUE)
```

Scalar `fData` columns become `obs`, matrix ones become `obsm`, the extra `assayData` elements
become layers, and everything parked on `experimentData@other` goes back into `uns` — so a
grassp user gets their object back with your results added. See `vignette("grasspio")`.

## Session info

```{r session, eval = TRUE}
sessionInfo()
```
